---
title: "Project 1"
author: "Kolton Wiebusch"
date: "9/15/2020"
output: 
      github_document:
        toc: true
---

```{r render, eval=FALSE, include=FALSE}
rmarkdown::render("Project_1.Rmd", output_file = "README.md")
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```
# Purpose
The purpose of this project is to create a vignette reading and summarizing data taken from NHL API's in order to analyze the data.

# Initial Setup
This first section goes through the behind the scenes work of creating functions to pull in data for later use

## Required Packages 
These are the packages used in this project.
```{r packages, message=FALSE}
library(tidyverse)
library(knitr)
library(jsonlite)
library(httr)
```

## NHL Records API  
This entails the creating of the functions to call on the NHL records data  
```{r endpoints1-2}
getFranchise <- function(ID = NULL, ...){
  franchise <- GET("https://records.nhl.com/site/api/franchise")
  franchise <- content(franchise, "text")
  franchise <- fromJSON(franchise, simplifyDataFrame = TRUE, flatten = TRUE)
  franchise <- as.data.frame(franchise)
    if(is.numeric(ID)){
      return(filter(franchise, data.mostRecentTeamId == ID))
    } else if(is.character(ID)){
      return(filter(franchise, data.teamCommonName == ID))
    }
  return(franchise)
} 

getFranchiseTotals <- function(ID = NULL, ...){
  totals <- GET("https://records.nhl.com/site/api/franchise-team-totals")
  totals <- content(totals, "text")
  totals <- fromJSON(totals, simplifyDataFrame = TRUE, flatten = TRUE)
  totals <- as.data.frame(totals)
    if(is.numeric(ID)){
      return(filter(totals, data.franchiseId == ID))
    } else if(is.character(ID)){
      return(filter(totals, data.teamName == ID))
    }
  return(totals)
}
```

```{r endpoint 3}

getFranchiseSeasonRecords <- function(ID = NULL, ...){
  if(is.character(ID)){
  recordsURL <- GET("https://records.nhl.com/site/api/franchise-season-records")
  records <- content(recordsURL, "text")
  records <- fromJSON(records, simplifyDataFrame = TRUE, flatten = TRUE)
  records <- as.data.frame(records)
  return(filter(records, data.franchiseName == ID))
  }else if(is.numeric(ID)){
  recordsURL <- GET(paste0("https://records.nhl.com/site/api/franchise-season-records?cayenneExp=franchiseId=", ID))
  records <- content(recordsURL, "text")
  records <- fromJSON(records, simplifyDataFrame = TRUE, flatten = TRUE)
  records <- as.data.frame(records)
  return(records)
  }
}
```

```{r endpoint 4}
getFranchiseGoalieRecords <- function(ID = NULL, ...){
    if(is.character(ID)){
  goalie <- GET("https://records.nhl.com/site/api/franchise-goalie-records")
  goalie <- content(goalie, "text")
  goalie <- fromJSON(goalie, simplifyDataFrame = TRUE, flatten = TRUE)
  goalie <- as.data.frame(goalie)
  return(filter(goalie, data.franchiseName == ID))
  }else if(is.numeric(ID)){
  goalie <- GET(paste0("https://records.nhl.com/site/api/franchise-goalie-records?cayenneExp=franchiseId=", ID))
  goalie <- content(goalie, "text")
  goalie <- fromJSON(goalie, simplifyDataFrame = TRUE, flatten = TRUE)
  goalie <- as.data.frame(goalie)
  return(goalie)
  }
}
```

```{r endpoint 5}
getFranchiseSkaterRecords <- function(ID = NULL, ...){
  if(is.character(ID)){
  skater <- GET("https://records.nhl.com/site/api/franchise-skater-records")
  skater <- content(skater, "text")
  skater <- fromJSON(skater, simplifyDataFrame = TRUE, flatten = TRUE)
  skater <- as.data.frame(skater)
  return(filter(skater, data.franchiseName == ID))
  }else if(is.numeric(ID)){
  skater <- GET(paste0("https://records.nhl.com/site/api/franchise-skater-records?cayenneExp=franchiseId=", ID))
  skater <- content(skater, "text")
  skater <- fromJSON(skater, simplifyDataFrame = TRUE, flatten = TRUE)
  skater <- as.data.frame(skater)
  return(skater)
  }
}
```

## NHL Stats API
This entails the creating of the function to call on the NHL stats data  
```{r statsEndpoint}
   #Setting up modifiers
  "?expand=team.roster" -> mfirst
  "?expand=person.names" -> msecond
  "?expand=team.schedule.next" -> mthird
  "?expand=team.schedule.previous" -> mfourth
  "?expand=team.stats" -> mfifth
  "?expand=team.roster&season=" -> msixth
  "?teamId=" -> mseventh
  "?stats=statsSingleSeasonPlayoffs" -> meighth
  
  #Creating function to get stats 
getNHLStats <- function(ID = NULL, modifier = NULL, ...){
    stats <- GET("https://statsapi.web.nhl.com/api/v1/teams")
    stats <- content(stats, "text")
    stats <- fromJSON(stats, simplifyDataFrame = TRUE, flatten = TRUE)
    stats <- as.data.frame(stats)
    if(is.numeric(ID) & is.character(modifier)){
      stats <- GET(paste0("https://statsapi.web.nhl.com/api/v1/teams/", ID, "/", modifier))
      stats <- content(stats, "text")
      stats <- fromJSON(stats, simplifyDataFrame = TRUE, flatten = TRUE)
      stats <- as.data.frame(stats)
      return(stats)
    }else if(is.character(ID) & is.character(modifier)){
      stats <- GET(paste0("https://statsapi.web.nhl.com/api/v1/teams/", ID, "/", modifier))
      stats <- content(stats, "text")
      stats <- fromJSON(stats, simplifyDataFrame = TRUE, flatten = TRUE)
      stats <- as.data.frame(stats)
      return(stats)
    }else if(is.numeric(ID)){
      return(filter(stats, teams.franchiseId == ID))
    } else if(is.character(ID)){
      return(filter(stats, teams.name == ID | teams.teamName == ID | teams.abbreviation == ID | teams.shortName == ID))
    }
  return(stats)
}
```

## Wrapper Function  
This function acts as a wrapper for all the previously discussed endpoints for NHL records and stats.  
```{r wrapper}
getNHL <- function(FUN, ID = NULL, m = NULL, ...){
  FUN <- match.fun(FUN)
  ID <- ID
  modifier <- m
  return(FUN(ID, modifier, ...))
}
```

# Exploratory Data Analysis
The next section's purpose is to explore and analyze the NHL data with specifications detailed in the assignment instructions.

## Joining of Datasets
Joining two data sets(according to franchiseAbbrv) for further data exploration
```{r join}
#Filtering for active franchises only and renaming to get ready for join
getFranchiseTotals() %>% filter(data.activeFranchise == 1) %>% rename(franchiseAbbrv = data.triCode) -> fTT

#narrowing down the stats data to return only a few variables
getNHLStats() %>% rename(franchiseAbbrv = teams.abbreviation) %>% select(franchiseAbbrv, teams.division.name, teams.conference.name) -> fTT2

#Full join on franchiseAbbrv
full_join(fTT, fTT2) -> joinedStats
```

## New variables
Creating new variables for later analysis
```{r newVars}
joinedStats %>% mutate(goalDifferential = data.goalsFor - data.goalsAgainst) -> joinedStats
joinedStats %>% mutate(penaltyMinsPerGame = data.penaltyMinutes/data.gamesPlayed) -> joinedStats
joinedStats %>% mutate(winPercentage = data.wins/data.gamesPlayed) -> joinedStats
joinedStats %>% mutate(shutoutPercentage = data.shutouts/data.wins) -> joinedStats
joinedStats %>% mutate(shootoutWinPercentage = ((data.shootoutWins)/(data.shootoutWins + data.shootoutLosses))) -> joinedStats
```

```{r separation}
#This code chunk separates joinedStats into Regular season and playoff stats for active teams
joinedStats$data.gameTypeId <- as.factor(joinedStats$data.gameTypeId)
levels(joinedStats$data.gameTypeId) <- c("RegularSeason", "Playoffs")
joinedStats %>% filter(data.gameTypeId == "RegularSeason" & teams.division.name != is.na(joinedStats$teams.division.name)) -> RegularSeasonStats
joinedStats %>% filter(data.gameTypeId == "Playoffs" & teams.division.name != is.na(joinedStats$teams.division.name)) -> PlayoffStats
```


## Contingency tables
Here, some contingency tables are returned. For the first one, you can see a breakdown of how many teams are in each division and conference overall. For the second and third tables, I wanted to take a closer look at my *favorite* team, the **St. Louis Blues**. The second table gives a breakdown of their current roster by position type, and the third table goes a step further and gives a breakdown of the number of position names within each position type. 
```{r contTables}
kable(table(RegularSeasonStats$teams.conference.name, RegularSeasonStats$teams.division.name))

#Return the Blues roster and break it down
getNHLStats(19, mfirst) -> g
as.data.frame(g[[31]]) -> BluesRoster
kable(table(BluesRoster$position.type))
kable(table(BluesRoster$position.name, BluesRoster$position.type))
```

## Numerical Summaries
Here are some numerical summaries on quantitative variables that I have created at different settings of categorical variables in the dataset. 

The first summary shows that teams amount more penalty minutes per game in the playoffs than in the regular season. When broken down further in the second summary, we can see that the Central division amounts considerably less penalty minutes per game than the other three divisions, and that the Pacific division not only averages the most penalty minutes per game in the playoffs, but also has the greatest jump in MPG from the regular season to the playoffs.  

The third summary shows averages and standard deviations for the shutout percentage, or the percent of time that the winning team does not allow the opposing team to score a goal all game. We can see a jump in percentages from the regular season to the playoffs, possibly because of more stifling defense. Also, the Eastern conference makes the better improvement between the two conferences between game types. The last summary breaks the conferences down into divisions. The Atlantic division in the Eastern conference seems to be the source of this jump, as their average shutout percentage in the playoffs is the highest among all groups, while the Pacific division has the lowest average with the highest standard deviation. 
```{r numSums}
#Filtering out teams that are not current at this moment
joinedStats %>% filter(teams.division.name != is.na(joinedStats$teams.division.name)) -> activeTeamStats

#Returning numerical summaries
activeTeamStats %>% group_by(data.gameTypeId) %>% summarise(avgPenaltyMPG = round(mean(penaltyMinsPerGame), 3), sdPenaltyMPG = round(sd(penaltyMinsPerGame), 3))

activeTeamStats %>% group_by(data.gameTypeId, teams.conference.name, teams.division.name) %>% summarise(avgPenaltyMPG = round(mean(penaltyMinsPerGame), 3), sdPenaltyMPG = round(sd(penaltyMinsPerGame), 3))

activeTeamStats %>% group_by(data.gameTypeId, teams.conference.name) %>% summarise(avgShutoutPctg = round(mean(shutoutPercentage), 3), sdShutoutPctg = round(sd(shutoutPercentage), 3))

activeTeamStats %>% group_by(data.gameTypeId, teams.conference.name, teams.division.name) %>% summarise(avgShutoutPctg = round(mean(shutoutPercentage), 3), sdShutoutPctg = round(sd(shutoutPercentage), 3))

```

## Plots  
This section includes various plots giving a visual summary of the data.

### Aggression Plot
These two scatter plots show how aggression(penalty minutes per game) correlates with winning(win percentage) in both the regular season and playoffs. There seems to be a slight negative correlation in the regular season, but flips to a slight positive correlation in the playoffs. In other words, it appears that there is a small indication that teams who are more aggressive and rough up their opponents have a better chance of winning in the playoffs. 
```{r aggression}
aggression1 <- ggplot(RegularSeasonStats, aes(x = penaltyMinsPerGame, y = winPercentage))
aggression1 + geom_text(aes(label = franchiseAbbrv))  + geom_text(x = 17, y = .55, size = 4, label = paste0("Correlation = ", round(cor(RegularSeasonStats$penaltyMinsPerGame, RegularSeasonStats$winPercentage), 2))) + ggtitle("Regular Season Win Percentage vs Penalty Minutes Per Game for All Teams")

aggression2 <- ggplot(PlayoffStats, aes(x = penaltyMinsPerGame, y = winPercentage))
aggression2 + geom_text(aes(label = franchiseAbbrv))  + geom_text(x = 20, y = .38, size = 4, label = paste0("Correlation = ", round(cor(PlayoffStats$penaltyMinsPerGame, PlayoffStats$winPercentage), 2))) + ggtitle("Playoff Win Percentage vs Penalty Minutes Per Game for All Teams")
```

### Division Goal Differentials Plot
This boxplot shows summaries of goal differential for each division in the NHL, playoff and regualar season combined. It appears that the Pacific division historically has the teams with the average worst goal differential, and has outliers that are the two lowest goal differential totals in the entire NHL. Oppositely, the Atlantic division has teams that appear to be historically dominant, with two teams that have a historic goal differential of about +2000 and +3500. 
```{r goalDifferentials}
goaldiffs <- ggplot(activeTeamStats, aes(x = teams.division.name, y = goalDifferential))
goaldiffs + geom_boxplot(fill = "blue") + xlab("NHL Divisions") + ggtitle("NHL Division Total Goal Differentials")
```





